Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-01-27 draft_lm1.tex
improved code to generate the table; added script to generate rt plot improved manuscript

Dengue is a disease which is caused by 4 different types of viruses 
which co-circulate between the vector and human populations. Immunity is 
considered permanent after the exposure of humans to each of the viruses and 
cross immunity is considered very limited[ref].
Thus, the proportions of viral types co-circulating at any point in time is 
strongly dependent on previous incidence of the disease, which determines the 
number of susceptibles for each viral type at any point in time.

Dengue transmission is modulated by environmental conditions, of which 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence~\cite{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient temperature variation dengue is predominantly a 
summer disease. So it is fair to say that these environmental fluctuations play 
a key role in determining beginning and end of epidemic periods through its 
effects on the force of infection.

@@

Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.\todo[inline]{anyone else?}
In the case of Dengue the 
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergenge of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

@@

\subsection{The dataset} 
The data used in this paper consists of  Dengue weekly notified cases time 
series for the city of Rio de Janeiro from 1996 to 2014. The cases are notified 
based only on clinical symptoms. Laboratory confirmation and serotype 
information are available only for a very small sample and only on recent 
years (2010-2013).

@@

A Susceptible-Infectious-Removed (SIR) model is proposed to model dengue 
dynamics.
In the traditional formulation of the model, transmission is governed by a 
transmission rate $\beta$ and recovery happens at a rate $\tau$.

As the epidemic progresses, the effective transmission  rate changes and is 
given by 
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number as defined 
in~\ref{eq:Rtestimate}.
We parameterize our model with the varying transmission using the system of 
ordinary differential equations
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
Of course, this is a rather simplified model, in which, for instance, the 
vector is omitted.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
can not discriminate between them due to the lack of serotype-specific data.
Nevertheless, this modeling strategy can still provide some insight into the 
disease dynamics and allows us to estimate the initial fraction of susceptibles 
$S_0$.

@@

The time series used for all the analyses consists of officially notified 
dengue cases aggregated weekly. 
Figure XX  shows the $R_t$ series, according to 
(\ref{eq:Rtestimate}). 
It can be seen that the inter-epidemic periods are characterized by $R_t$ being 
indistinguishable from 1.

Figure \ref{Fig:S0}, shows the model from (\ref{eq:model}) fitted to the data. 
In it we can see that the susceptibles series in each epidemic starts at the 
estimated level of $S_0$.
The proportion of susceptibles may seem low, but we 
must remember that these estimates are being affected by an unknown 
underreporting factor, which experts suggest is somewhere between 5 and 10, 
i.e. for every case observed there are 5 or 10 unobserved.
Since this 
underreporting ration affects both the numerator and denominator of 
(\ref{eq:AR}), its effects should cancel out, giving us an unbiased attack ratio 
estimate.
One other possible source of bias which would lead to the 
underestimation of $S_0$ could come from a significant part of the population 
not being exposed to the disease.
But, as we can see in figure xx\todo{add 
figure with the four maps}, the disease spreads pretty much homogeneously over 
the entire city (in the last four epidemics, at least).

@@

Table \ref{tab:AR} contains the attack ratios calculated for each 
epidemic/outbreak.
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for $R(t)$.
The authors are also grateful to Claudia T. 
Codeço for helpful discussions about the manuscript.

\bibliography{lm1}
\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}
\begin{figure}
\begin{center}
\includegraphics[width=.9\textwidth]{plots/concat_SI.png}
\end{center}
\caption{
{\bf Susceptibles and Infectious posterior curves.}  The curves were estimated 
only for the periods where $R_t> 1$.  The susceptible curves in the top panel 
reflect the prevalence of fraction of susceptibles to circulating strain(s) for 
each epidemic/outbreak. In the lower panel see the posterior distribution 
of infectious curves, represented by its Median and 95\% credible interval. 
Scatter plot shows the observed cases. Prevalences are scaled as fractions 
of the entire population.
}
\label{Fig:S0}
\end{figure}

@@

\begin{table}[!ht]
\caption{
\bf{Median Attack ration calculated according to (\ref{eq:AR2})}}
\begin{center}
\begin{tabular}{l|r}
\hline
Year & Median Attack Ratio (\%) \\
\hline
1996 & \textbf{38.64} (17.18-54.19)\\
1997 & \textbf{87.31} (73.83-87.31)\\
1998 & \textbf{49.86} (49.19-49.86)\\
1999 & \textbf{10.76} (3.72-20.24)\\
2000 & \textbf{25.49} (24.47-26.98)\\
2001 & \textbf{48.12} (46.57-49.15)\\
2005 & \textbf{14.74} (10.23-21.38)\\
2006 & \textbf{11.22} (8.00-14.41)\\
2007 & \textbf{15.08} (14.76-15.38)\\
2008 & \textbf{13.73} (3.10-30.84)\\
2010 & \textbf{18.30} (17.34-19.33)\\
2011 & \textbf{8.59} (8.20-9.36)\\
2012 & \textbf{13.92} (12.71-14.92)\\
\hline
\end{tabular}
\begin{flushleft} We report medians and 95\% credibility intervals.
\end{flushleft}

>>>>

Dengue is a disease which is caused by 4 different types of viruses 
which co-circulate between the vector and human populations. Immunity is 
considered permanent after the exposure of humans to each of the viruses and 
cross immunity is considered very limited\cite{halstead_dengue_2007}.
Thus, the proportions of viral types co-circulating at any point in time is 
strongly dependent on previous incidence of the disease, which determines the 
number of susceptibles for each viral type at any point in time.

Dengue transmission is modulated by environmental conditions, of which 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence~\cite{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient temperature variation dengue is predominantly a 
summer disease. So it is fair to say that these environmental fluctuations play 
a key role in determining beginning and end of epidemic periods through its 
effects on the force of infection, which cannot be taken as 
constant\cite{reiner_time-varying_2014}

@@

Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.
In the case of Dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergenge of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

@@

\subsection{The dataset} 
The data used in this paper consists of  Dengue weekly notified cases time 
series for the city of Rio de Janeiro from 1996 to 2014. The cases are notified 
based only on clinical symptoms. Laboratory confirmation and serotype 
information are available only for a very small sample and only on recent 
years (2010-2013). The incidence is normalized for the parameter estimation 
procedures. The normalizatio was done by dividing the number os cases reported 
by the total city population at each year as given by the census.

@@

A Susceptible-Infectious-Removed (SIR) model is proposed to model dengue 
dynamics.
In the traditional formulation of the model, transmission is governed by a 
constant transmission rate $\beta$ and recovery happens at a rate $\tau$.

For our analysis we chose to let the force of infection vary with time, just 
as it does in the actual epidemics, as seen in the data. So as the epidemic 
progresses, the effective transmission  rate changes and is 
given by 
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number, estimated as 
in~\ref{eq:Rtestimate}.
The complete model with the time-verying force of infection is given by
the system of ordinary differential equations below.
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
Of course, this is a rather simplified model, in which, for instance, the 
vector is omitted.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
can not discriminate between them due to the lack of serotype-specific data.
Nevertheless, this modeling strategy can still provide some insight into the 
disease dynamics and allows us to estimate the initial fraction of susceptibles 
$S_0$.

@@

Figure \ref{fig:rtseries}  shows the $R_t$ series, according to 
(\ref{eq:Rtestimate}). 
It can be seen that the inter-epidemic periods are characterized by $R_t$ being 
indistinguishable from 1.

\begin{figure}
 \centering
 \includegraphics[width=16cm]{./plots/rt_series.png}

 \caption{{\bf Estimated time-series for $R_t$, along with 95\% credible 
intervals.} Top panel show reported cases from which $R_t$ is estimated.}
 \label{fig:rtseries}
\end{figure}

Figure \ref{Fig:S0}, shows the model from (\ref{eq:model}) fitted to the data. 
In it we can see that the susceptibles series in each epidemic starts at the 
estimated level of $S_0$.
The proportion of susceptibles may seem low, but we 
must remember that these estimates are being affected by an unknown 
underreporting factor, which experts suggest is somewhere between 5 and 10, 
i.e. for every case observed there are 5 or 10 unobserved.
Since this 
underreporting ration affects both the numerator and denominator of 
(\ref{eq:AR}), its effects should cancel out, giving us an unbiased attack ratio 
estimate.
One other possible source of bias which would lead to the 
underestimation of $S_0$ could come from a significant part of the population 
not being exposed to the disease.
However, as we can see in figure \ref{fig:mapas} , the disease spreads pretty 
much homogeneously over 
the entire city (in the last four epidemics, at least).

@@

Table \ref{tab:AR} contains the attack ratios calculated for each 
epidemic/outbreak. It is interesting to notice that the larger epidemics, in 
terms of total number of cases are not the one with the greater attack ratios. 
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for $R(t)$.
The authors are also grateful to Claudia T. 
Codeço for helpful discussions about the manuscript.

\bibliography{lm1}
\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}
\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{plots/concat_SI.png}
\end{center}
\caption{
{\bf Susceptibles and Infectious posterior curves.}  The curves were estimated 
only for the periods where $R_t> 1$.  The susceptible curves in the top panel 
reflect the prevalence of fraction of susceptibles to circulating strain(s) for 
each epidemic/outbreak. In the lower panel see the posterior distribution 
of infectious curves, represented by its Median and 95\% credible interval. 
Scatter plot shows the observed cases. Prevalences are scaled as fractions 
of the entire population.
}
\label{Fig:S0}
\end{figure}

@@

\begin{table}[!ht]
\caption{
\bf{Median Attack ration calculated according to (\ref{eq:AR2})}. $^\dag$: Year 
corresponds to the start of the epidemic. Sometimes the peak of cases occur in 
the following year.}
\begin{center}
\begin{tabular}{l|r}
\hline
Year$^\dag$ & Median Attack Ratio (\%) \\
\hline
1996 & \textbf{38.64} (17.18-54.19)\\
1997 & \textbf{87.31} (73.83-87.31)\\
1998 & \textbf{49.86} (49.19-49.86)\\
1999 & \textbf{10.76} (3.72-20.24)\\
2000 & \textbf{25.49} (24.47-26.98)\\
2001 & \textbf{48.12} (46.57-49.15)\\
2005 & \textbf{14.74} (10.23-21.38)\\
2006 & \textbf{11.22} (8.00-14.41)\\
2007 & \textbf{15.08} (14.76-15.38)\\
2008 & \textbf{13.73} (3.10-30.84)\\
2010 & \textbf{18.30} (17.34-19.33)\\
2011 & \textbf{8.59} (8.20-9.36)\\
2012 & \textbf{13.92} (12.71-14.92)\\
\hline
\end{tabular}
\begin{flushleft} We report medians and 95\% credibility intervals.
\end{flushleft}
